\documentclass[12pt]{article}

\title{Notes on Zermelo's axioms for set theory}

\author{Randall Holmes}

\usepackage{amssymb}

\begin{document}

\maketitle

I'm going to organize these point by point using the section numbers in Zermelo's paper, with some intermissions.

\begin{enumerate}

\item  He introduces a domain {\bf B} of all objects.  He says that some of these objects are {\em sets\/} (primitive notion).  He introduces equality as a primitive notion.

\item  In section 2, he introduces membership $a \in b$ as a primitive relation.  He tells us that if $a \in b$, $b$ must be a set (and telegraphs that there can be a set with no elements, but there can be only one:  this follows from a later axiom).  In any event what he says here is enough to see that if there is any object which is not a set, it has no elements.

It is a feature of most modern treatments of foundations of mathematics based on set theory that all objects are sets.  This isn't a natural assumption philosophically, though philosophers have gamely played along with it, and it is nice to see that in this source text this assumption is not made.

\item In section 3, he defines the subset relation:  $M \subseteq N$ iff $M$ and $N$ are sets (important) and for all $x$, if $x$ is an element of $M$ then $x$ is an element of $N$.  Notice that for any $x$ which is not a set and any $y$, it is true that any $z$ which belongs to $x$ (there aren't any) belongs to $y$, but we do not say $x \subseteq y$.  He also defines disjointness of sets.

\item In section 4, he introduces the idea of an assertion $\phi$ being ``definite", meaning that it can be decided whether it is true or false.  This is important to him, and more in a philosophical than a mathematical way.  A propositional function $P(x)$ of an object $x$ ranging over a class $K$ is definite if $P(x)$ is definite for each element of the class $K$.  Question:  what does he mean by ``class"?  Is {\bf B} a class in this sense?  He does know that {\bf B} is not a set (see section 10).  He tells us that $a \in b$ is always definite, and that $M \subseteq N$ is always definite
(because the propositional function $x \in N$ of $x$ is definite for each $x \in M$, I interpret him as saying).

He now introduces some axioms:

\begin{description}

\item[Axiom I:]  If $M \subseteq N$ and $N \subseteq M$ then $M=N$.  Sets with the same elements are the same.  This is called the axiom of extensionality.

\item[informal definition:]  He introduces the notation $\{a,b,c,\ldots,r\}$ for a finite set whose elements are exactly $a,b,c,\ldots,r$.  A motto of mine:  whenever a mathematician introduces those rows of dots, he is cheating.

\item[Axiom II:]  There is a set 0 with no elements (why does he feel constrained to call it ``fictitious"?);  for any object $a$ the set $\{a\}$ which has $a$ as an element and no other element exists;  for any two objects $a,b$, the set $\{a,b\}$ which has $a$ as an element, has $b$ as an element and has no other elements exists.  This is called the axiom of elementary sets.

\end{description}

\item In section 5 he observes that the sets 0, $\{a\}$ and $\{a,b\}$ are uniquely determined by axiom I.  He also notes that $a=b$ is always ``definite" because it is equivalent to
$a \in \{b\}$.  I was going to say in protest that it is definite because it is equivalent to $a \subseteq b \wedge b \subseteq a$, but that is actually inadequate:  that only shows that equality of sets is definite.  The question of why Zermelo wants to talk about definiteness of assertions is really interesting.  Does he ever say anything that isn't definite?

The notation $\emptyset$ for the empty set is more usual now, but we will stick with 0.

\item In section 6 he raises a really interesting issue in philosophy of set theory as it were in passing.  He notes that $0 \subseteq M$ and $M \subseteq M$ are always true:  he defines a ``part" of a set $M$ as a subset of $M$ other than 0 or $M$.  We might think of an object as a part of itself, and prefer to say ``proper part" here.  We can feel with Zermelo discomfort with the idea that all sets have a common part 0.  But in any event, the philosophical point to make is that the elements of a set are not parts of the set (whatever a part of a set may be taken to be).  The relation of part to whole, however it is defined, must be transitive, and $a \in \{a,b\}$, $\{a,b\} \in \{\{a,b\}\}$ but $a \not\in \{\{a,b\}\}$.  A set does not have its elements as its parts but its subsets;  it does have (or can be understood as having) its singleton subsets as atomic parts correlated with its elements.  David Lewis wrote a whole book about this, {\em Parts of Classes\/}, which I quite recommend as reading.

Now (still in section 6) we get

\begin{description}

\item[Axiom III:]  Whenever the propositional function $P(x)$ is definite for each element of a set $M$, there is $M_P$ such that $M_P \subseteq M$ and for every $x$,
$x \in M_P$ if and only if $x \in M$ and $P(x)$.  This is called the Axiom of Separation.

We introduce the notation $\{x \in M:P(x)\}$ for the set $M_P$.

This axiom gives the kind of ability to define objects correlated with properties which Frege wanted in his Axiom V governing ``courses of values".  But the restriction of
this formation of objects from properties to properties of elements of a previously given set seems to make this workable without contradiction.  Moreover, nothing is being given up
in terms of actual mathematical practice:  we do not construct sets of mathematical interest by considering properties of all objects taken indiscriminately, but by considering properties of objects of a particular sort.

Zermelo talks about the importance of ensuring that the property $P(x)$ is ``definite".  I am quite interested in what he thinks this notion is doing for him.

\end{description}

\item  In section 7, Zermelo notes that for $M_1 \subseteq M$ (and in fact for any $M_1$) we can define $M \setminus M_1$ as $\{x \in M: x \not\in M_1\}$,  Zermelo calls this the complement of $M_1$;  it would more usually be called the complement of $M_1$ relative to $M$.  This is provided by Axiom III.

\item In section 8, Zermelo introduces the intersection $[M,N]$ of sets $M$ and $N$, provided by axiom III as $\{x \in M:x \in N\}$.  This is more usually written $M \cap N$ now.  The ability to write the assertion that $M$ and $N$ are disjoint as $[M,N]=0$ is worth noting.

\item In section 9 he extends intersections first to intersections $[M,N,R, \ldots]$ of a list of given sets and then to the notion $\bigcap T$ of the intersection of all elements of
a nonempty set $T$.  His analysis of this is interesting.  He says that by Axiom III for a given set $T$ and each $a$ we can define a subset $T_a = \{t \in T:a \in t\}$ of all elements of $T$
which contain $a$.  It is then a definite question for each $a$ whether $T_a=T$ (that is, whether every element of $T$ contains $a$) and so by Axiom III there is a
set $\bigcap T = \{a \in t:T_a=T\}$, where $t$ is any element of $T$ (it does not matter which one you use).  A more usual definition would be $\{a \in t:(\forall u \in T:a \in u)\}$:  but Zermelo in his formulation avoids using a quantifier.

\item In section 10 we see the form which the Russell paradox argument takes in this theory.  It does not lead to paradox.

\begin{description}
\item[Theorem:]  For each set $M$, there is a set $M_0$ such that $M_0 \subseteq M$ and $M_0 \not\in M$.

\item[Proof:]  It is definite for each $x \in M$ whether $x \in x$ (and Zermelo takes pains to note that nothing in his system prevents $x \in x$ from being true for some $x$).

Let $M_0 = \{x \in M:x \not\in x\}$.  Clearly $M_0 \subseteq M$.

Now either $M_0 \in M_0$ or not.  If $M_0 \in M_0$ then $M_0$ contains an $x$ such that $x \in x$, which is incompatible with its definition.  So $M_0 \not\in M_0$.
But then $M_0 \in M$ is impossible, as if we had $M_0 \in M$ it would meet the conditions to belong to $M_0$, and we have already shown that it cannot meet these condiitions.
So $M_0 \not\in M$ as desired.

\end{description}

Zermelo then observes that it follows from this that we cannot have a set which contains every object, and so (interestingly) he observes that the domain {\bf B} cannot be a set.  It is philosophically very interesting that he makes this observation:  one could further ask what sort of thing {\bf B} is$\ldots$ and this is a natural opening to the later idea of proper classes.

Still in section 10, Zermelo introduces two more axioms.

\begin{description}

\item[Axiom IV:]  To every set $A$ there corresponds a set ${\cal P}(A)$, the power set of $A$, whose elements are exactly the subsets of $A$.  (Axiom of power set)

\item[Axiom V:]  To every set $A$, there corresponds a set $\bigcup A$, the union of $A$, whose elements are exactly the elements of the elements of $A$.  (Axiom of union)

\item[side comments of ours about these axioms:]  It is very interesting that the union $\bigcup A$ has to be provided by axiom where the related set $\bigcap A$ is provided already by Axiom III.

A technical point about union is that $\bigcup \{A\} = A$ if $A$ is a set:  if we made the technical modification that $\bigcup \{x\} = x$ where $x$ is a non-set, we could use
$\bigcup$ in lieu of a definite description operator.  This is also a technical advantage of supposing that everything is a set.   It would also be reasonable to introduce notation
$\Theta$ with the properties $\Theta(\{x\}) = x$ for all $x$, and $\Theta(x) = 0$ if $x$ is not a singleton.

An interesting point about the power set, which is related to the intellectual origins of Russell's paradox and the theorem of section 10, is the theorem of Cantor that the
power set of $A$, even for $A$ infinite, is larger in size than the set $A$.

We give an informal argument for this.  We say that two sets $A$ and $B$ are the same size if there is a bijection from $A$ to $B$, that is, a function $F$ from $A$ to $B$ which is one to one and onto $B$.  The set ${\cal P}(A)$ is at least as large as $A$ (each element $a$ of $A$ corresponds to $\{a\} \in {\cal P}(A)$).  Suppose that $A$ is at least as large as ${\cal P}(A)$, that is, there is a one-to-one map $f$ from ${\cal P}(A)$ to a subset of $A$.  This map has an inverse $f^{-1}$ defined for some but not all elements of $A$.  For the sake of argument we extend $f^{-1}(a)$ to be 0 for $a$ in $A$ which is not in the range of $f$.  Now define $R$ as $\{a \in A:a \not\in f^{-1}(a)\}$.  We ask whether $f(R)$, a well-defined element of $A$ by hypothesis, is an element of $R$:  this will be true if and only if $f(R) \in A$ and $f(R) \not\in f^{-1}(f(R))$.  Now $f(R) \in A$ is true and $f^{-1}(f(R))= R$, so we have
$f(R)$ an element of $R$ if and only if $f(R)$ is not an element of $R$, which is a contradiction:  there can be no such $f$.

This is not a proof in Zermelo's system, because we do not yet know how to talk about functions like $f$ in Zermelo's system, but shortly we will.

\end{description}

\item In section 11 we define binary unions of sets.  We define $M+N$ (usually now written $M \cup N$) as $\bigcup \{M,N\}$.  More generally we define $M+N+R +\ldots$
as $\bigcup \{M,N,R,\ldots\}$.  Some algebraic laws such as $M+0=M+M=M$ are noted.

\item In section 12 it is noted that the commutative and associative laws $M+N=M+N$ and $(M+N)+R = M+(N+R)$ hold for union of sets.  Further, distributive laws
$[M+N,R] = [M,R]+[N,R]$ and $[M,N] + R = [M+R,N+R]$ hold.  Zermelo notes that these theorems are proved by Axiom I and logic.

\item If $M$ is a set different from 0 and $a \in M$, it is definite whether $M=\{a\}$.  It is definite whether a set has one element or not.  This is Zermelo's language:  here is my own argument:  for any set $M$. it is definite whether there is an element $a$ of $M$ such that every element $x$ of $M$ is equal to $a$.  If there is such an element $a$ of $M$, then $M=\{a\}$, and if there is not, $M$ is not a singleton.

We say that a set $T$ is a pairwise disjoint collection of sets or a partition if for each pair $A,B$ of distinct elements of $T$ we have that $A$ and $B$ are disjoint.

Let $T$ be a pairwise disjoint collection of sets.  We define $\prod T$ as the collection of all subsets $X$ of $\bigcup T$ such that for each $A \in T$, $A\cap X$ has exactly one element.
A product $\prod\{M,N\}$ is written $MN$ and a product $\prod\{M,N,R,\ldots\}$ is written $MNR\ldots$.

Before stating the next axiom, it is worth noting the relationship between the addition and multiplication operations now defined on sets and addition and multiplication of numbers.
If $M$ has $m$ elements and $N$ has $n$ elements and $M,N$ are disjoint, then $M+N = M \cup N$ has $m+n$ elements and $MN$ has $mn$ elements.  Similar statements are true for sums and products of finite collections of disjoint sets.

We would like it to be the case that a product of infinitely many nonempty sets has to be nonempty, and that is what the next axiom says.  The final axiom provides us with something no previous axiom has done, an example of an infinite set.

\begin{description}

\item[Axiom VI:]  If $T$ is a pairwise disjoint collection of sets and $0 \not\in T$, then $\prod T$ is nonempty.  In other words, for any pairwise disjoint collection $P$ of nonempty sets, we can find a set which contains exactly one element of $P$:  we can choose one element from each set and collect them.  This is called the Axiom of Choice.

{\bf A Holmes side remark:}  the Axiom of Choice allows us to choose an element from each element of a possible infinite pairwise disjoint set $T$.  An evil fact is that Zermelo's theory allows us to choose for each element $t$ of a set $T$ a set {\em not\/} in $T$ without using Choice and collect the resulting objects into a set. The collection of all sets $t_0 = \{x \in t:x \not\in x\}$ does not belong to $t$ and does
belong to the power set of $t$ and so to the power set of $\bigcup T$.  So we can define the ``negative choice set" $N_T$ as $\{u \in {\cal P}(\bigcup T):(\exists t \in T:u = \{x \in t:t \not\in t\})\}$:  we can choose an element of the power set of the union of $T$ which does not belong to $t$ for each $t \in T$ and collect these into a set, without using Axiom VI.  If $T$
is a partition (a pairwise disjoint collection of nonempty sets) and contains no $t$ which is its own singleton set, then a distinct non-element will be chosen for each element of $T$.  Of course, there is less to this than it seems:  if we have $(\forall x:x \not\in x)$, then $t_0 = t$ and $N_T = T$.  But it is still fun.

\item[Axiom VII:]  There is a set $Z$ such that $0 \in Z$ and for each element $a$ of $Z$, $\{a\}$ is also an element of $Z$. (axiom of infinity).

Observe that Axiom II already gives us infinitely many distinct objects, $0, \{0\}, \{\{0\}\},\ldots$, but nothing before axiom VII allows us to construct an infinite set.

\item[what if everything is a finite set?]  Nothing under this heading is in Zermelo:  these are entirely my modern comments.

I actually pause to investigate this.  I give an alternative collection of axioms:

\begin{description}

\item[A:]  same as Axiom I:  sets with the same elements are the same.

\item[B:]  The empty set 0 exists.  For each object $a$, $\{a\}$ exists.

\item[C:]  for each nonempty set $A$ and object $b$, $A + \{b\} = A \cup \{b\}$ exists.

We note that it would work simply to assert that the empty set 0 exists and for any set $x$ and object $y$, $x \cup \{y\}$ exists.

\item[D:]  for each propositional function $P(x)$ such that $P(0)$, for each object $a$, $P(a) \rightarrow P(\{a\})$, and for each set $A$ and object $b$,
$P(A) \wedge P(b) \rightarrow P(A \cup \{b\})$ is true, we have that $P(x)$ is true for all $x$.  What Axiom D is saying is that we allow only those objects which
are either 0, or the singleton of a previously constructed object, or of the form $A \cup \{b\}$ where $A$ and $b$ are already constructed:  we are only allowing sets built by finite listing from previously given sets.

In line with the suggestion under Axiom C, this could be simplified to say, for any propositional function $P(x)$, that if (1) $P(0)$ is true and (2) for any set $x$ and object $y$,
if $P(x)$ and $P(y)$ then we have $P(x \cup \{y\})$, then we have (3) for any $z$ we have $P(z)$.  It should also be noted that it is an immediate consequence of Axiom D that every object is a set.

\end{description}

Axioms A,B,C should be recognizable as consequences of Axioms I, II, and V.  Axiom D should remind one of the principle of mathematical induction.

Though we do not have a formal definition of the notion of a finite set, you should notice that if the propositional function $P(x)$ is ``$x$ is a finite set",
the conditions of axiom D apply, so we should expect that everything is actually a finite set in the system described by axioms $A-D$.

Now the punchline is that axioms I-VI (but not axiom VII) are all consequences of axioms A-D.

\begin{enumerate}

\item[I:] is easy:  it is the same as axiom A.

\item[II:]  axiom B says that 0 and $\{a\}$ exist;  axioms B and C give us $\{a\}+\{b\} = \{a,b\}$.

\item[III:] Let $Q(x)$ be a propositional function.  We show using axiom D that the predicate $P(x)$ asserting that $\{y \in x:Q(y)\}$ exists holds of every x.

$P(0)$ is true because $\{y \in 0:Q(y)\} = 0$.

$P(\{a\})$ is true (whether or not $P(a)$ is true) because $\{y \in \{a\}:Q(y)\}$ is either 0 or $\{a\}$ depending on whether $Q(a)$ is true,
and both of these sets exist.

Suppose $P(A)$ is true.  Then whether or not $P(b)$ is true $\{y \in A + \{b\}:Q(y)\}$ exists because it is either $\{y \in A:Q(y)\}$ (if $\neg Q(b)$) which
exists by hypothesis, or it is $\{y \in A:Q(y)\} \cup \{b\}$, if $Q(b)$ is true, which exists by hypothesis and axiom C.

Thus Axiom III (for the specific propositional function $Q(x)$, but we made no special assumptions about it, so the argument is completely general) holds.

\item[V:]  Fix a set $a$.  $a \cup 0 = a$ exists.  $a \cup \{b\}$ exists by axiom C.  $a \cup (b \cup \{x\}) = (a\cup b) \cup \{x\}$ exists by axiom C if $a \cup b$ exists.  So by axiom D
$a \cup x$ exists for any set $x$.  

We use this to prove the existence of general unions of sets:
$\bigcup 0=0$.  $\bigcup \{a\} = a$.  $\bigcup (A \cup \{b\}) = (\bigcup A) \cup b$, and we have shown that binary unions exist just above, so by Axiom D $\bigcup x$ exists for every $x$.

\item[IV:]  We show the existence of the set $\{y \cup \{x\}:y \in A\}$ for any set $A$ and object $x$.  $\{y \cup \{x\}:y \in 0\} = 0$.
$\{y \cup \{x\}:y \in \{u\}\} = \{u \cup \{x\}\}$ which exists by application of Axiom C followed by Axiom B.  Suppose that $\{y \cup \{x\}:y \in A\}$ exists. Then $\{y \cup \{x\}:y \in A \cup \{b\}\}$ is the union of
$\{y \cup \{x\}:y \in A\} \cup \{b \cup \{x\}\}$, which exists by hypothesis that the set exists for $A$ and applications of axioms B and C.  By application of axiom D, $\{y \cup \{x\}:y \in A\}$ exists for every $A$.

${\cal P}(0) = \{0\}$ which exists.  ${\cal P}(\{a\}) = \{\{a\}\} \cup \{0\}$, which exists.  Suppose ${\cal P}(A)$ exists.  ${\cal P}(A \cup \{x\})$  is the union of ${\cal P}(A)$ and the set $\{a \cup \{x\}:a \in {\cal P}(A)\}$, which we just showed to exist.  So by Axiom D every set has a power set.

\item[VI:]  We want to prove that if $P$ is a partition, $P$ has a choice set.  If $P=0$, 0 is a choice set for $P$.  If $P= \{a\}$, $a$ is a nonempty set with an element $x$
and $\{x\}$ is a choice set for $\{a\}$.  If $P = A + \{b\}$ is a partition and $A$ has a choice set $C$, then $b$ is a nonempty set with an element $x$ and $C \cup \{x\}$ is a choice set for $A+\{b\}$.  By Axiom D, every partition has a choice set.

\item[VII:]  We argue that no set can have its elementwise image under the singleton operation as a proper subset.  This shows that Axiom VII is false if Axioms A-D hold, because
an inductive set will have its elementwise image under singleton as a proper subset.
The exact property $P(x)$ which we consider is the property that no subset of $x$ has its elementwise image under the subset operation as a proper subset.   

0 has this property (its elementwise image under the singleton map is 0, which is not a proper subset of 0, and it has no other subsets).

$\{a\}$ cannot have its elementwise image under the singleton operation as a proper subset, because its only proper subset is 0, and its image under the singleton operation is the nonempty set $\{\{a\}\}$.  The only other subset of $\{a\}$ is the empty set already covered.  

Now suppose that $A \cup \{b\}$ has a subset $B$ whose image under the elementwise application of the singleton operation is a proper subset of $B$, and that $A$ does not have this property.  Let $B$ be a subset of $A \cup \{b\}$ which has its elementwise image under the singleton map as a proper subset.  Clearly $b \in B$, as otherwise we would have a subset of $A$ with this property contrary to hypothesis.  Clearly $b$ is the singleton of some $c \in A$, or $B \setminus \{b\}$ would have the property under consideration.  We now consider the intersection $C$ of all sets which contain $\{b\}$ and are closed under singleton ($B$ is such a set, so we can construct $C$ as the intersection of all subsets of $B$ with this property).  If $C$ does not contain $b$ as an element, it is a subset of $A$ which has a proper subset as its elementwise image under the singleton operation.  if $C$ does contain $b$ as an element, then $C$ is its own image under the singleton operation (the set of singletons of elements of $C$ would contain $\{b\}$ and be closed under the singleton operation) so $B \setminus C$ would have its own elementwise image under the singleton map as a proper subset:  $B \setminus C$ certainly includes its elementwise image under the singleton map in this case, and it must include it properly or else the singleton image of $B$ would be all of $B$, contrary to hypothesis.  In this argument we are able to make fancy use of Axiom III because we have shown that Axiom III follows from Axioms A-D.



\end{enumerate}



So in fact all the axioms I-VI hold (and axiom VII does not) if we allow only the construction of finite sets by listing.

We do not want to restrict ourselves in this way, since we do want to consider things like the collection of natural numbers and the collection of real numbers, and their general subsets.

\item[another incidental Holmes remark about Axiom VII:]  Some pathologies of Zermelo set theory from a modern standpoint would be repaired by formulating Axiom VII in a more general way: 

\begin{description}

\item[Axiom VII$^*$:]  Let $F(x)$ be any object-valued function (any formal definition of an object which gives an object $F(x)$ for each object $x$ in a well-defined way), and let $a$ be any object.  Then there is a set
$I_F$ such that $a \in I_F$ and for each $x \in I_F$ we have $F(x) \in I_F$. 

 Zermelo's axiom is of course the instance of this with $F(x) = \{x\}$.\end{description}

We could then define an $F,a$-inductive set as a set $X$ such that $a \in X$ and for each $x \in X$ we have $F(x) \in X$.  And we can then define $I_F^0$, or informally $\{F^n(a):n \in {\mathbb N}\}$, as
the intersection of all $F,a$-inductive elements of ${\cal P}(I_F)$.

The theory with this axiom is somewhat stronger, though not nearly as strong as ZFC, and it allows construction of sets such as $V_{\omega}$ and the transitive closure of a general set which Zermelo set theory does not allow one to define.   The axiom in this form could be called the Axiom of Iteration.

The Axiom of Replacement which is usually adjoined to Zermelo to address such concerns, and which is far stronger, can be stated in the same style:  for any object-valued function
$F(x)$ (meaning simply a well-defined expression which picks out an object for each $x$) and any set $A$, the set $F``A$ defined as $\{F(a):a \in A\}$ (the set of all $F(a)$ for $a \in A$) exists.  This looks innocent but is amazingly strong.  It is worth noticing that if the graph of $y = F(x)$ is given as a propositional function $G(x,y)$ of two variables, $F(x)$ can
be defined in terms of $G(x,y)$ as $F(x) = \bigcup \{y:G(x,y)\}$, under the further condition that $G(x,y)$ implies that $y$ is a set.  My reasons for noting this are similar to Frege's reasons for introducing his $\backslash$ operator. $\{y:G(x,y)\}$ will presumably exist, since the fact that $G$ is a graph implies that for any given $x$, $G(x,y)$ is true of no more than one object.

The Axiom of Replacement introduces a whole new very strong idea;  the Axiom of Iteration simply generalizes what Zermelo was doing with his axiom of infinity.  The use of Frege-style object valued functions as well as propositional functions in the background pleases me.

\end{description}

\item In section 14 we define the actual set of natural numbers, which we might call $\mathbb N$ and which Zermelo calls $Z_0$.

We define an {\em inductive set\/} as a set which contains 0 and for each $a$ contains $\{a\}$ if it contains $a$.  So Axiom VII simply asserts the existence of
an inductive set.

That a set is inductive is a definite property.

There is not necessarily a set of all inductive sets, but by Axiom III there is a set $I_Z$ of all inductive elements of ${\cal P}(Z)$.  Note that $Z \in I_Z$, so $I_Z$ is
nonempty.  Define $Z_0$, or $\mathbb N$, as $\bigcap I_Z$.  It should then be observed that for any inductive set $Z'$, whether it is a subset of $Z$ or not,
$Z \cap Z'$ is inductive and belongs to $I_Z$, and so $Z_0 \subseteq Z'\cap Z \subseteq Z$, from which we see that $Z_0$ is the intersection of {\em all\/} inductive sets.



We informally define a function from natural numbers (whatever they may be)  to their representatives in Zermelo's theory:  define $\# 0$ as the set $0$ and recursively define
$\# (n+1)$ as $\{\# n\}$.   If being an image under $\#$ is a definite predicate, it is clearly inductive.  Further, by ordinary mathematical induction, it has no inductive proper subsets, so it actually is $Z_0$.  The final step is to suggest that the answer to what the natural numbers are might be that they {\em are\/} the Zermelo natural numbers (the elements of $Z_0$).


\item [remark:]  This ends the axiomatics part of the paper.  What follows is the theory of equivalence, basically foundations of the theory of infinite cardinal number.

\item  Two disjoint sets $M$ and $N$ are said to be {\em immediately equivalent\/}, $M \sim N$, if there is a subset $\Phi$ of their product $MN$ such that each element of
$M+N = M\cup N$ occurs as an element in one and only one element of $\Phi$.

Such an element $\Phi$ is called a mappinng of $M$ onto $N$ (notice that this is a symmetric notion, unlike our usual treatment).

If $\{m,n\} \in \Phi$ we say that each element is mapped to or corresponds to the other.

It should be clear that existence of $\Phi$ gives us a one to one correspondence between elements of $M$ and elements of $N$:  each element $m$ of $M$ belongs to
exactly one element of $\Phi$, which itself contains a unique element of $N$, which we might by an abuse of notation write $\Phi`m$.  Similarly, each element of $N$ is associated with a unique $\Phi`n$.  This notation is a new idea, I will see if it is useful in my exposition.  It is useful to note that $\Phi`\Phi`x=x$ for each $x \in M+N$.

$\Phi`x = \bigcup \{y \in \bigcup \Phi:y \neq x \wedge \{x,y\} \in \Phi\}$, or $$\Phi`x = \Theta \{y \in \bigcup \Phi:y \neq x \wedge \{x,y\} \in \Phi\}$$ if we introduce this notation.  Using union notation presumes that the elements of $M+N$ are sets or presumes the modification of the definition of union proposed above.

Now if we have an informal one to one correspondence $f:M \rightarrow N$, we can define $\Phi$ as $\{\{x,f(x\}\in MN: x \in M\}$:  this requires that $f$ be definite in a suitable sense.

\item  He begins this section by observing that it is definite for any disjoint sets $M$ and $N$ and subset $\Phi$ of $M+N$ and $x \in M+N$ whether the set of elements of $\Phi$ that contain $x$ has one element.  Thus it is also evident whether all elements of $M+N$ have this property, that is, whether $\Phi$ is a mapping from $M$ to $N$.  Thus by Axiom III
we can define a subset of ${\cal P}(M+N)$ containing exactly the mappings from $M$ to $N$, and it is definite whether this set is 0 or not.  Thus it is definite for sets $M,N$ whether
$M \sim N$.  This section is entirely about definiteness and should be read closely for Zermelo's intentions.  What interests me is whether he ever considers a quantifier over all
of {\bf B}, or over some proper class, to be definite.

\item If $\Phi$ is a mapping from $M$ onto $N$ then each subset $M_1$ of $M$ is mapped to a subset $N_1$ of $N$ by a subset $\Phi_1$ of $\Phi$.  The set $\Phi_1$ can be defined
as $\{x \in \Phi: x \cap M_1 \neq 0\}$.   The set $N_1$ can be defined as $N \cap \bigcup \Phi$.  Now each element of $M_1$ occurs in only one element of $M_1$, because it occurs in at least one by definition of $\Phi_1$, and if it appeared in more than one it would also appear in more than one element of $\Phi$.  Similarly, an element of $N_1$ appears in at least one element of $\Phi_1$ by definition of $\Phi_!$ and in at most one because it occurs in at most one element of $\Phi$.

On our own hook, we introduce the notation $\Phi\lceil M_1$ for $\Phi_1$ and $\Phi``M_1$ for $N_1$.

\item If $M$ is disjoint from $N$ and $M \sim R$ and $N \sim R$ then $M \sim N$.   Similarly, if $M \sim R$, $R \sim R'$ and $R' \sim N$, with $M$ and $N$ disjoint, then
$M \sim N$.

If $\Phi$ is a mapping from $M$ onto $R$ and $\Psi$ is a mapping from $R$ onto $N$, we define $\Phi \circ \Psi$ as $\{\{x,\Psi`(\Phi`(x))\} \in MN:x \in M\}$.  Here my notation helps.
It should be clear that $\{\{x,\Phi`(\Psi`(x))\} \in MN:x \in N\}$ is exactly the same set, and a mapping from $M$ onto $N$. 

We define the composition of three mappings in the same way and draw the same conclusions.

\item  \begin{description}

\item[Theorem:]  If $M$ and $N$ are any two sets, there is always a set $M'$ which is equivalent to $M$ and disjoint from $N$.

\item[Proof:]  Let $S = \bigcup(M+N)$ and let $r = \{u \in M+S:u \not\in u\}$, for which we know by the theorem of section 10 that $r \not\in M+S$.  Then the sets $M$ and
$R = \{r\}$ are disjoint, and the product $M' = MR$ has the property required to witness the theorem.

Suppose that some $\{m,r\} \in MR$ belongs to $N$.  Then $r \in \bigcup N \subseteq M+S$, because $\bigcup N \subseteq \bigcup (M+N)=S$.  And this is impossible
because $r \not\in M+S$.  This verifies that $M'$ is disjoint from $N$.

That $M \sim M' = MR$ is witnessed by the mapping $$\{\{m,\{m,r\}\} \in MM':m \in M\}.$$  We do need to note that $M$ and $M'$ are disjoint, because
if some $\{m,r\}\in M$ we would have $r \in \bigcup M \subseteq M+S$, because $\bigcup M \subseteq \bigcup (M+N) = S$.

\item[Corollary:]  He notes specifically here that it is thus shown that no set $T$ can contain all sets equivalent to a nonempty set $M$, because for any set $T$ there is a set equivalent to $M$ and
disjoint from $\bigcup T$ and so (because nonempty) not an element of $T$.  This makes the point that Frege's (or Cantor's) definition of cardinal number is not compatible with this theory of sets.  It is historically significant that he makes this point.

\item[An obvious observation:]  It is fascinating that Zermelo here makes explicit practical use of the Russell paradox diagonalization.

\item[A further Holmes observation:]  This implies further, though Zermelo does not draw this conclusion, that we have a general representation of all functions from $M$ to $N$.
Let $F(x)$ be any function such that for any $x \in M$ we have $F(x) \in N$.  Let $r$ be defined as above.  Then we can encode the function $F$ by the set
$\{\{\{m,r\},F(m)\}\in M'N:x \in M\}$.  For any $x \in M$, there is just one element of $f$ such that $\{\{m,r\},n\} \in N$ (there is no converse result for elements of $N$).
We can define $f`_{M\rightarrow N}(x)$ as $\bigcup \{y \in N: x\in M \wedge \{\{x,r\},y\} \in N\}$.  Notice that we arrange for this to be 0 if $x \not\in M$, and that the apparent extra parameter $r$ is actually exactly determined by $M$ and $N$.

Implicit in this is a definition of a peculiar Cartesian product $M \times N = \{\{\{x,r\},y\}\in M'N:x \in M \wedge y \in N\}$ and a definition of set representations of functions in basically the modern way.  The definition of the ordered pair suggested here is weird because it depends on the intended domain and codomain of the function, but this might even be thought to have virtues.

Just for laughs, a complete suite of definitions:

We define $\Delta(M,N)$ as $\{u \in M + \bigcup(M+N):u \not\in u\}$.

We define $M \times N$ as $$\{\{\{x,\Delta(M,N)\},y\}\in (M\{\Delta(M,N)\})N :x \in M \wedge y \in N\}.$$

We define $(\lambda x \in M \rightarrow N:F(x))$ as $$\{\{\{x,\Delta(M,N)\},y\} \in M \times N:y = F(x)\}.$$

We define $f`_{M\rightarrow N}(x)$ as $$\bigcup \{y \in N: x\in M \wedge \{\{x,\Delta(M,N)\},y\} \in N\}.$$

or $$\Theta \{y \in N: x\in M \wedge \{\{x,\Delta(M,N)\},y\} \in N\},$$  if we introduce the $\Theta$ notation.

This gives a complete suite of definitions for construction and application of functions whose values are sets, coded as sets.  It will work for all functions
with the technical modification of $\bigcup A$ which generalizes the equation $\bigcup \{x\}=x$ to non-sets $x$, or if we use $\Theta$.

\end{description}

\item  In section 20, Zermelo wants to verify that it is always definite for sets $M$ and $N$ whether there is a set $R$ disjoint from both $M$ and $N$ and equivalent to both $M$ and $N$.

The question of whether the specific set $M'$ defined in section 19 is equivalent to $N$ is a definite question, clearly.  And this is equivalent to the question under consideration:
if $M \sim R \sim N$ with $M$ disjoint from $R$ and $N$ disjoint from $R$, then $M \sim M' \sim N$ similarly.  That $M \sim M'$ is shown in section 19.  $M' \sim M \sim R \sim N$ implies
$M' \sim N$ by the second result of section 18.

\item  We extend the definition of $M \sim N$ to the case where $M$ and $N$ are not disjoint:  we define $M \sim N$ as holding when there is $R$ disjoint from $M$ and $N$ such that
$M \sim R \sim N$.  We say that $M$ is mediately equivalent to $N$.  

Zermelo says that the equivalence of $M$ and $N$ is witnessed by two mappings, $\Phi$ from $M$ onto $R$ and $\Psi$ from $R$ onto $N$.  An element of $m$ then corresponds to $\Psi`\Phi`m \in N$, and an element $n$ of $N$ corresponds to $\Phi`\Psi`n \in m$, this being a one-to-one correspondence in the usual sense.

He also notes that to each subset $M_1$ of $M$ there corresponds a subset $N_1$ of $N$ defined as $\Psi``\Phi``M_1$, which is equivalent to $N_1$ by application of the methods of section 17 (with $R_1 = \Phi``R$ as the mediating set).

It is definite whether two sets are equivalent in this way.

\item  In this section, Zermelo argues that equivalence as extended in section 21 is reflexive and transitive (one can note that it is clearly symmetric).

$M \sim M$ is witnessed by using section 19 with $M=N$ to build a set disjoint from $M$ and equivalent to $M$.

Suppose $M\sim R$ and $N \sim R$.  We then have for some $M'$ and $N'$ disjoint from $M, R$ and $N,R$ respectively that $M \sim M' \sim R$ and $N \sim N' \sim R$.
Use section 19 to get a set $R'$ disjoint from $M+N+R$ and equivalent to $R$.  We then have $M \sim M' \sim  R \sim R'$ and so $M \sim R'$ by section 18, and
similarly $N \sim N' \sim  R \sim R'$ and so $N \sim R'$ by section 18, and so $M \sim N$ by the extended definition with $R'$ as the mediating set.

\item  The null set 0 is equivalent only to itself.  Singletons $\{a\}$ are equivalent to all other singletons $\{b\}$ and to no other sets.

The first point is made by observing that $0M = 0$ and so no set other than 0 can be immediately equivalent to 0, and so no set other than 0 can be mediately equivalent to 0.

If $\{a\}$ is disjoint from $M$ and $\Phi$ is a mapping from $\{a\}$ onto $M$, then there is exactly one element $\{a,b\}$ of $\Phi$ containing $a$,
and any element of $\Phi$ must contain an element of $\{a\}$ and so contain $a$ and so be $\{a,b\}$.  Now $b \in M$ follows, and further for any element $c$ of $M$,
$c$ must belong to some element of $\Phi$ and so to the element $\{a,b\}$ of $\Phi$, so $c=b$, so $M = \{b\}$ which is a singleton.  And of course $\{\{a,b\}\}$ is a mapping from $\{a\}$ onto $\{b\}$ for any $a,b$.

\item  The theorem of this section amounts to a proof that addition of cardinal numbers is definable.

If $M \cap N = 0$ and $M' \cap N'=0$ and $M\sim M'$ and $N\sim N'$ then $M+N \sim M'+N'$.

The size of a union of two disjoint sets is determined by the sizes of the two disjoint sets individually, in effect.

If $M+N$ is disjoint from $M'+N'$ then if $\Phi$ is a mapping from $M$ onto $M'$ and $\Psi$ is a mapping from $N$ onto $N'$ then
$\Phi + \Psi$ is a mapping from $M+N$ to $M'+N'$.  Each element of $\Phi + \Psi$ is a pair $\{p,q\}$ with $p \in M+N$ and $q\in M'+N'$ (because it is either a pair with
$p \in M$ and $q \in N$ or it is a pair with $p\in M'$ and $q \in N'$).  For each $p \in M+N$ there is exacty one $q \in M'+N'$ such that $\{p,q\} \in \Phi + \Psi$ and vice versa by symmetry.  This is because $p \in M+N$ belongs to exactly one of $M$ and $N$, and in the first case we find a unique $q$ such that $\{p,q\} \in \Phi$ and no element of $\Psi$ which contains $p$,
and in the second case we find a unique $q$ such that $\{p,q\} \in \Psi$, and no element of $\Phi$ which contains $p$.

If $M+N$ is not disjoint from $M'+N'$, find a set $M''$ equivalent to $M$ and disjoint from $M+N+M'+N'$ and a set $N''$ equivalent to $N$ and disjoint from $M+N+M'+N'$.
Then we have $M+N \sim M''+N''$ by the previous paragraph and $M''+N'' \sim M'+N'$ by the previous paragraph, and we then have mediate equivalence of $M+N$ and $M'+N'$.  This is not how Zermelo did it;  it is noticeably simpler.

\item This is a lemma supporting the Schr\"oder-Bernstein theorem.

\begin{description}

\item[Theorem:]  If a set $M$ is equivalent to one of its parts $M'$, then if $M \subseteq M_1 \subseteq M'$ then $M \sim M_1$.

Notice that this is a situation which can only hold if $M$ is infinite:  no finite set is equivalent to one of its (proper) parts.

\item[Proof:]  Let $M \sim M' \subseteq M_1 \subseteq M$ and $Q = M_1-M'$.  There is $M''$ such that $M \sim M'' \sim M' \subseteq M_1 \subseteq M$, $\Phi$ is a mapping from
$M$ onto $M''$ and $\Psi$ is a mapping of $M''$ onto $M'$.  For any $A \subseteq M$ we have an image $A' =\Psi``\Phi`A \subseteq M'$.  We can define $$I = \{A \in {\cal P}(M):\Psi``\Phi``A \subseteq A \wedge Q \subseteq A\}.$$  I is nonempty because $M \in I$, so we can define the intersectioin $\cap I$ of all elements of $I$, which is in effect the set
of all iterated images of elements of $Q$ under application of the mapping $\Psi \circ \Phi$.

Note that $Q \subseteq \bigcap I$, since all elements of $I$ have $Q$ as a subset.  $\Psi``\Phi``\bigcap I \subseteq \bigcap I$ should be evident.  In fact $\bigcap I \in I$.
Further, $\bigcap I = Q + \Psi``\Phi``\bigcup I$.  The set of objects which are either in Q or the image under the mapping of some element of $\Psi``\Phi``\bigcup I$ belongs to $I$:  it includes $Q$ and it has the right closure properties:  anything which is an image under the mapping of an element of $Q$ must be in any element of $I$, and anything which is
an image under the mapping of an element of the intersection of all elements of $I$ is in any element of $I$.  And any element of this set must belong to $\bigcap I$, as already shown.

Now $M_1 = Q+\Psi``\Phi``M = (Q + \Psi``\Phi``\bigcap I) + (\Psi``\Phi``M - \Psi``\Phi``\bigcap I) = \bigcap I \cup  (\Psi``\Phi``M - \Psi``\Phi``\bigcap I)$,
a union of two disjoint sets.

Now $\bigcap I \sim \Psi``\Phi``\bigcap I$ and $(\Psi``\Phi``M - \Psi``\Phi``\bigcap I)$ is equivalent to itself, so $M_1$ is equivalent to the union of
$\Psi``\Phi``\bigcap I$  and $(\Psi``\Phi``M - \Psi``\Phi``\bigcap I)$, by the theorem of the previous section, that is, to $\Phi``\Psi``M = M'$, which is in turn equivalent to $M$.



\end{description}

\item  In this section Zermelo proves that if a set $M$ is equivalent to one of its proper parts, then it is equivalent to any set $M \cup \{x\}$ or $M - \{x\}$ obtained by adding
or removing a single element.

If $M$ is equivalent to $M' \subseteq M$ and $x \in M - M'$, then $M' \subseteq M - \{x\} \subseteq M$ and $M' \sim M$ implies $M - \{x\} \sim M$ by the theorem of the previous section.

If $x\in M$ and $y \not\in M$ then $M -\{x\} \cup \{y\} \sim M$.  Let $M''$ be equivalent to $M$ via a mapping $\Phi$ and disjoint from $M \cup \{x\} \cup \{y\}$.
Then $\Phi \lceil (M - \{x\}) \cup \{y,\Phi`x\}$ is a mapping from $M''$ to $M -\{x\}\cup \{y\}$, so $M$ is mediately equivalent to $M -\{x\} \cup \{y\}$.

Now if $M$ is equivalent to $M' \subseteq M$ and $x \in M'$, choose $y \in M-M'$ and we have $M \sim M' \sim M'-\{x\}\cup \{y\} \subseteq M-\{x\} \subseteq M$, and so
in this case as well $M-\{x\} \sim M$.

Now suppose that $M' \subseteq M$ and $M \neq M'$ and $M \sim M'$ and $x \not\in M$.  We then have $M' \cup \{x\} \subseteq M \cup \{x\}$ and $M' \cup \{x\} \neq M \cup \{x\}$.
If we have $M \cup \{x\} \sim M'\cup \{x\}$ we then have $M = (M \cup \{x\}) - \{x\} \sim M \cup \{x\}$ by the previous results of this section:  but we do have this equivalence
by the theorem on equivalences of unions of disjoint sets.

I don't think I did this quite the same way Zermelo did, but it is similar.

\item This section proves the Schr\"oder-Bernstein theorem.

\begin{description}

\item[Theorem:]  If $M \sim M' \subseteq N$ and $N \sim N'\subseteq M$ then $M \sim N$:  if each of the sets $M$ and $N$ is equivalent to a subset of the other then
$M$ and $N$ are equivalent.

\item[Proof:]  The subset $N'$ of $M$ has in its turn a subset $M''$ equivalent to $M'$ and so to $M$, simply because $M'$ is a subset of $N$ and $N$ is equivalent to $N'$.
So $M'' \subseteq N' \subseteq M$ and $M \sim M''$ implies $N' \sim M$ and so $M \sim N$.



\end{description}


\item \begin{description}

\item[Theorem:]Let $T$ be a collection of sets and $Z$ a given set.  We can define a set $T' = \{F(A):A \in T\}$ for a function $F$ which we present with the property that  $A \sim F(A)$ for
each $A \in T$ and
for distinct $A,B \in T$ we have $F(A)$ and $F(B)$ disjoint from one another and from $Z$.

\item[Proof:]  Let $S = \bigcup T$.  Let $T''$ be chosen equivalent to $T$ and disjoint from $T \cup S \cup \bigcup (S+Z)$ by the method of section 19.
We thus have a mapping $\Phi$ from $T$ onto $T''$.

For each $A \in T$, the set $F(A)$ is $\{\{x,\Phi`A\} \in ST'':x \in A\}$.  $F(A)$ is disjoint from $F(B)$ for $A \neq B$ because $\{x,\Phi`A\} = \{y,\Phi`B\}$ would require
either $x = \Phi`B$ or $y = \Phi`A$, which is impossible because $T''$ is disjoint from $S$.  $F(A)$ is disjoint from $Z$ because $T''$ is disjoint from the union of $Z$.
$A \sim F(A)$ is witnessed by the set of pairs $\{x,\{x,\Phi`A\}\}$ in $A\cdot F(A)$, which is a pair because $T''$ is disjoint from the union of $S$ ( no element of $T''$ can be an element of the element $x$ of $A \subseteq S$).

Each $F(A)$ belongs to ${\cal P}(ST'')$ so there is a set $$\{F(A) \in {\cal P}(ST''):A \in T\}.$$ with the desired properties.





\end{description}


\item  Zermelo proves a general principle of choice for collections of sets which are not necessarily pairwise disjoint.

The statement of his theorem is ``{\em General principle of choice.\/}  If $T$ is a set whose elements $\ldots$ are all sets different from the null set, there always exist sets $P$
that, according to a certain rule, uniquely correlate with each element $M$ of $T$ one element $m$ of that $M$."

Let $T$ be a set of nonempty sets.

Construct by the method of the previous section a set $X=\{F(A):A \in T\}$ such that $A \sim F(A)$ for each $A \in T$ and $F(A)$ and $F(B)$ are disjoint 
for distinct $A$ and $B$.

Now use Axiom VI to construct a set $P$ which contains exactly one element of each element of $X$.  Then we have a precise rule to choose an element
from each element $A$ of $T$:  take the unique element of $P \cap F(A)$, which will be a pair of the form $\{x,\Phi`A\}$ (using the notation of the proof in the previous section)
and the element $x$ will be an element of $A$.

What is very interesting about this is that Zermelo asserts the existence of a rule without saying what the rule is as an object.  We know from above that in fact
he can present a set coding this or any function from a set $M$ to a set $N$ (in this case $M$ would be the set $T$ and $N$ would be $\bigcup T$).  But it appears that he is in effect proving a statement that says a certain second order entity (a function from objects to objects) exists.  The hypothesis of his proof of the Well-Ordering Theorem has the same second order quality.  He does say that $P$, a set, codes this correspondence:  this could be tidied up a bit with further references to the mapping $\Phi$, which also plays an essential role.  I would use the method of my comment on section 19 to simply code an explicit choice function, thus requiring no references to second order entities at all.

The exact definition (using our notation introduced in our section 19 notes) is $P' = (\lambda A \in T \rightarrow \bigcup T:\bigcup (\bigcup(P \cap F(A))\cap A))$.  Technically this only works if $T$ is a set of sets of sets, or if we recast it as $P' = (\lambda A \in T \rightarrow \bigcup T:\Theta (\bigcup(P \cap F(A))\cap A))$.





\end{enumerate}


\end{document}