\documentclass[12pt]{article}

\title{Document Inventory for The Loglan Institute}

\author{Randall Holmes}

\begin{document}

\maketitle

\begin{description}

\item[Books (oldest to most recent):]

\begin{enumerate}

\item  Loglan 1, 1975 hardcover -- 2 copies

\item  Loglan 4 and 5, 1975 hardcover -- 2 copies

\item  Loglan 1, 1975, trade paperback, 2 copies

\item Loglan 4 and 5, 1975, trade paperback, 3 copies

\item  Loglan 1, 1989, trade paperback, 2 copies.  Of these documents, only this one is ``official", but the 1975 edition is of serious interest as a statement of intent.   See statement below under NB3.

\end{enumerate}

\item[Reports (most recent to oldest):]

\begin{description}

\item[NB3:]   Notebook 3, The Present State of the Loglan Language, 1987:  an original and a photocopy.  This has also been scanned and OCR'ed.  It is the official view of the Institute that nothing prior to NB3 is ``official":  NB3, the latest dictionaries, and Loglan '89 (with appendix H in the online version) give as complete a picture of the language as it came to us as we can get.

\item[NB1 and NB2:]  I have one copy of these two documents in a green folder, and some somewhat defective copies of this material from the Faith Rich Papers in Chicago.  This material must be photocopied and scanned.  It is not authoritative, but provides a valuable window into the history of the project.



\begin{enumerate}

\item Notebook 1:  the Machine Grammar and Corpus of Loglan, 1982

\item Notebook 2:  A Proposed Revision in the Structure of Loglan Words, 1982

\end{enumerate}

\item[Loglan 3:  1-3  Learning Loglan:]  1978, 1980 A small booklet and a manuscript of all or part of it.

\end{description}

\item[Periodicals:]

\begin{description}

\item

\item[The Loglanist (originally La Loglentan, later La Logli)]:

I have complete holdings of this, the main periodical of the Institute.

\begin{description}


\item[La Loglentan]:

\begin{enumerate}

\item  The Loglanist 1 (from Loglan 2, Methods of Construction, JCB 1970).  a manuscript.

\item The Loglanist 1 no 1 Aug 76 two bound copies

\item The Loglanist 1, no 2, Dec 76 2 copies

\item The Loglanist vol 1 no 3 Feb 1977 two copies

\item The Loglanist vol 1 no 4  Apr 1977 two copies

\item The Loglanist vol 1 no 5  Jun 1977 two copies

\item The Loglanist vol 1 no 6  Aug 1977 two copies

\item The Loglanist, volume 2, the Feb 1978 distribution.  manuscript, one bound copy.

\item The Loglanist, volume 2, the Jun 1978 distribution.  manuscript, one bound copy

\item vol 3 Feb 1979

\item vol 3 March 1979

\item vol. 3 October 1979

\item vol 3 Dec 1979

\item vol. 4 Jun 1980

\item  vol 4 August 1980

\item vol 4 Nov 1980 special issue:  a suplement to Loglan 1!  by John Parks-Clifford.  2 copies

 \item vol 4 Dec 1980

\item vol. 5 Feb 1981 (it says 1980 but actual date seems clear)

\item vol 5 Sept 1981  2 copies

\item vol 5 Dec 1981 2 copies

\item vol 6 Oct 1983  A special issue on morphology 2 copies

\item vol 7 April 1983  The first of two special issues on grammar -- the second never appeared.  2 copies

 \end{enumerate}

\item[La Logli]:  


\begin{enumerate}

\item  La Logli 96/1   contains Loglan 1 updater, other articles  3 copies

\item La Logli, 97/1, 97/2, 99/1 (three copies of each)  Loglan 3, Understanding Loglan, by Stephen L. Rice, in three installments.

\item La Logli, 02/1 two copies  First Visit to Loglandia, a novel, by Alex Leith

\end{enumerate}

\end{description}

\item [Lognet]:

newsletters as opposed to substantial magazines from various periods

My holdings for the last Lognet series are incomplete:  I indicate my lacks by comparing with the web page.  Much of what I dont have can be seen on the official Loglan page {\tt www.loglan.org}.

\begin{enumerate}
\item  Le Logla Cmata, Dec 1978.  I think there may only ever have been one issue.
\item Lognet 87 March 87 no 1

\item Lognet 87 June 1987 no 2

\item Lognet 88 May 1988 no 1.  The editorial promises some discussion of the Split.

\item Lognet 89/1 2 copies

\item I do not have 90/1 or 90/2

\item Lognet 90/3

\item I do not have any of the four issues from 1991

\item  I do not have any of the three issues from 1992

\item I do not have 93/1 or 93/3 or 93/4.

\item Lognet 93/2

\item Lognet 94/1

\item Lognet 94/2

\item I do not have 94/3

\item Lognet 95/1
\item Lognet 95/2

\item Lognet 96/1 2 copies

\item Lognet 96/2 two copies

\item Lognet 96/3

\item Lognet 97/1 two copies

\item Lognet 97/2 two copies

\item Lognet 98/1

\item Lognet 99/1 two copies

\item Lognet 00/1 two copies












\end{enumerate}


\end{description}

\item[Miscellaneous]:

I may add more comments about these items (mostly correspondence).

\begin{enumerate}

\item  A puzzle:  some handwritten pages of Loglan examples that I do not understand the purpose or context of.

\item 27 Oct 77 letter to Loglanist subscribers

\item 31 Jan 78 letter to Loglanist subscribers

\item Fall 78 letter to membership (grant not renewed)

\item 1 Dec 78 announcing Le Cmata which seems to have happened once

\item Loglan Board meeting 26 Feb 82

\item letter to membership 28 May 1985 announces Lognet

\item letter to membership 20 Jan 86

\item card announcing publication of NB3 Apr 86

\item letter to membership Nov 86  working on NB3

\item Aficionado Agreement re secrecy of NB3, 1987

\item 2 Aug 87, announcement of NB3 and the schism

\item letter to Jim Gillogly 15 May 1988

\item Midsummer Bulletin 1988

\item Autumn Bulletin 1988 some interesting discussion of Lojban incidentally

\item 1 May 1989 letter from chairman about release of confidentiality agreement about NB3 and request to extend it in part

\item  1 May 1989 announcement of Loglan 1 '89

\item Autumn 1989 Bulletin

\item 24 June 1996 letter to former buyers of books promoting the current periodicals

















\end{enumerate}


\end{description}


\end{document}