\documentclass[12pt]{article}

\title{Math 387 Homework 1}

\author{Randall Holmes}

\usepackage{amssymb}

\begin{document}

\maketitle

\begin{enumerate}

\item  Seven teams send their baseball teams to a tournament in which each time must play each other exactly once.
Find the answer in two different ways (same thing I did in problem 1).

\newpage

\item  You can have a hamburger, a fishburger or a turkey burger.  You can have a plain bun or a sesame seed bun.
There are four toppings:  bacon bits, onions, ketchup, and mustard.  You may have two toppings, which must be different.

How many kinds of burger are possible?

\newpage

\item  Explain why the ordered pair $(a,b)$ cannot simply be defined as the two element set $\{a,b\}$.

\newpage

\item  $A = \{1,2,3,4\}$ and $B = \{5,6,7\}$.

\begin{enumerate}

\item  How many functions from $A$ to $B$ are there?

\item How many functions from $B$ to $A$ are there?

\item How many functions from $A$ to $B$ are injections (one-to-one)?

\item How many functions from $B$ to $A$ are injections (one-to-one)?

\item How many functions from $A$ to $B$ are surjective (onto)?  This may be quite hard, depending on how you approach it.

\item How many functions from $B$ to $A$ are surjective (onto)?

\end{enumerate}

\newpage

\item   If six flavors of ice cream are available, and you are making a triple cone, 


\begin{enumerate}
\item and the order in which the scoops are placed makes a difference, how many cones are possible?  \item If order matters and the flavors all have to be different, how many cones are possible?  \item If order does not matter and the flavors all have to be different, how many cones are possible?  \item If order doesnt matter and flavors can be repeated, how many cones are possible?
\end{enumerate}

\newpage  

\item How many partitions are there of the set $\{1,2,3,4,5,\}$?   We worked this out for a four element set in class.





\end{enumerate}

\end{document}