\documentclass[12pt]{article}

\usepackage{amssymb}

\title{Math 189 Fall 2023 Homework 4}

\author{Dr Holmes}

\begin{document}
\maketitle

These are modelled on exercises in Levin, 0.3, and you should read Levin section 0.3 and my set theory notes for background.  One piece of notation which I did not use in lecture appears here:  if $A$ is a set, we use $|A|$ to represent the number of elements in $A$ (which may also be called the size or cardinality of $A$), if $A$ is a finite set.  I may provide you with references to similar problems in Levin's 0.3 set;  I also suggest doing the interactive examples in section 0.3, which are quite nice.

\begin{enumerate}

\item (modelled on problem 1 in the 0.3 exercises)

Let $A = \{1,2,4,7\}$ and let $B = \{1,2,4,8,16\}$.  Find each of the following sets and present them in list notation.

\begin{enumerate}

\item $A \cup B$

\item $A \cap B$

\item $A \setminus B$  [Levin uses $A \setminus B$ instead of $A - B$, so I will as well.  Expect the notes to be revised.]

\item $B \setminus A$

\end{enumerate}

\item  Find $|A|$ (the number of elements in $A$) for each of the following examples.  If the set is infinite, say so.  Modelled on problem 3.

\begin{enumerate}

\item  $\{17,18,19,\ldots,35\}$ [be careful that you don't make a fencepost error].

\item   $\{x^3:x \in \mathbb Z^+ \wedge x^3<200\}$ [read carefully, this is a set of perfect cubes and it is not very large]

\item  $\{1,2,3,4,5,6,7\} \times \{1,3,6,9,27,81\}$  List a few elements of this set (they are ordered pairs) and tell me how many elements the set has.  You don't want to list them all!

\item   $\{x \in \mathbb Z^+: x|60\}$

\item $\{x \in \mathbb Z^+:60|x\}$

\end{enumerate}

\item  This is problem 10 in Levin 0.3.  Let $A = \{x \in \mathbb N:3 \leq x \leq 13\}$, $B = \{x \in \mathbb N: x$ is even$\}$ and let $C = \{x \in \mathbb N: x$ is odd$\}$.  Find the following sets.  Write list or set builder notation for the sets which does not mention $A, B,$ or $C$.

\begin{enumerate}

\item   $A \cap B$

\item $A \cup B$

\item $B \cap C$

\item $B \cup C$


\end{enumerate}

\item Let $A = \{1,2,3\}$.  Write ${\cal P}(A)$ in list notation.

\item modelled on question 15.  Draw Venn diagrams representing each of the following sets.

\begin{enumerate}

\item $A \setminus (B \cup C)$

\item $(A \setminus B) \cup (A \setminus C)$

\item $A \setminus (B \cap C)$

\end{enumerate}

\item  Present sets $A$, $B$ of small positive integers in list notation such that $|A|=3, |B| = 4,$ and $|A \cup B| = 5$.  (remember that
$|X|$ means the number of elements in $X$ when $X$ is a set).

\item (extra credit puzzle question)  Explain why no set $A$ exists such that $A = \{2,|A|\}$.  Give me two different sets
$B$ such that $B = \{1,2,|B|\}$.  The first part of this question is number 29 in 0.3.

\end{enumerate}


\end{document}

