\documentclass[12pt]{article}

\title{Math 189, fall 2024, Homework 5}

\author{Randall Holmes}

\begin{document}

\maketitle

This is due Wednesday, Oct 2, after the exam, but the material is on Test I, so be sure you attack it promptly.

Give both exact answers computed with your calculator and a description of how they are computed.

I would in addition work problems from sections 1.1, 1.2 and 1.3, completing each problem before looking at the solution.

\begin{enumerate}

\item  A certain state has license plates consisting of three letters follow by five digits.

\begin{enumerate}

\item How many possible plates are there if there are no additional restrictions?

\item  How many possible plates are there if no letter or digit is repeated?

\item  How many plates have exactly one 8 on them? 

\item  How many plates have at least one 8 on them (one or more 8's?)

\end{enumerate}

\item  How many divisors does 1800 have?

\item Of 25 children, 11 like chocolate ice cream, 11 like vanilla ice cream, and 11 like strawberry ice cream.4 like chocolate and vanilla.  6 like chocolate and strawberry.  2 like vanilla and strawberry.  1 likes all three.   How many of the children do not like any of the flavors?  How many of the children will only eat strawberry ice cream?

\item  Expand $(x+y)^8$.

\item  Using the digits 3 through 9, find the number of 4 digit numbers such that

\begin{enumerate}
\item no additional condition are imposed (digits may be repeated freely and may appear in any order)

\item  No digit can appear more than once but they may appear in any order

\item  The digits must be distinct and must appear in increasing order

\end{enumerate}


\item  A committee of 20 members wants to choose a subcommittee with six members.  How many choices can be made if

\begin{enumerate}
\item  There are no additional instructions (just choose six of the twenty people)


\item  How many ways are there to choose the subcommittee of six members and in addition choose a chair and secretary from the subcommittee?

\item  There are ten men and ten women in the group of 20, and you are required to choose three men and three women to be on the subcommittee.

\end{enumerate}

\item   How many different anagrams of PARALLEL are there?  (A good math word)




\end{enumerate}

\end{document}